An enhancer code compendium of early human neural and neural crest development

### 2025-04-25 manuscript.tex
Abstract

%%% The summary (abstract) should consist of a single paragraph of 150 words or fewer.

%Context
Enhancers encode cell types as clusters of transcription factor binding sites. 
%need
Enhancer code studies have been limited to specific cell types in model organisms.
%task and results
We generate a single-cell multiome atlas of human neural tube organoids 
and of the head of a four weeks post conception human embryo. 
Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that 
are shared between the organoid model and the human embryo and we provide a compendium of enhancer codes 
for each individual cell type. 
We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 transcription factor binding sites in a nucleosome depleted region, using soft syntactic rules. 
We also provide counter examples to this rule, with enhancers encoded as homotypic clusters and by dimer motifs. 
Finally, we validate DeepNeuralTube using enhancer reporter assays and on human genetic variants.
%conclusion
The enhancer compendium is available as a resource at: something.aertslab.org
%implication

>>>>

By encoding clusters of transcription factor (TF) binding sites, genomic enhancers act as hubs for combinatorial TF
recruitment. 
Enhancer code studies have been limited to specific cell types in model organisms. 
Here we set out to create a compendium of enhancer codes for all cell types of the developing human neural tube and neural crest. 
To this end, we generated a single-cell multiome atlas of human neural tube organoids and of the head of a four weeks post conception human embryo. 
Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that are shared between the organoid 
model and the human embryo and comprehensively analyse the enhancer code of each individual cell type. 
We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 TF binding sites in a nucleosome depleted region, using soft syntactic rules. 
We also provide counter examples to this rule, with enhancers encoded as homotypic clusters and by dimer motifs. 
Finally, we validate DeepNeuralTube using enhancer reporter assays and assess the role of human genetic variants on 
facial mesenchyme enhancers. 
Our neural tube enhancer compendium is available as a resource at: something.aertslab.org
